---
title: "RVCOP Session: Hypothesis Testing in R"
author: "John Wainaina <phomwangi@gmail.com>"
date: "2023-05-26"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)


```

# Hypothesis testing in R

```{r eval=FALSE}

A statistical hypothesis is an assumption made by the researcher about the 
data of the population collected for any experiment.

To validate a hypothesis, it will use random samples from a population. 
On the basis of the result from testing over the sample data, it either
*selects/fail to reject*/**!accept!** or *rejects* the hypothesis.

A statistical test is a procedure used to evaluate the evidence 
the data provides against a hypothesis.

Statistical tests are either parametric or non-parametric.

**Parametric tests** - based on assumptions about the distribution 
of population from which the sample was taken. 

**Nonparametric statistics** - do not rely on any distribution. 
They are not based on any assumptions about distribution 
of the population in reference.

**Parametric tests** - continuous data/outcome

Mean is the most appropriate summary measure
for continuous outcome provided the data is not highly skewed.

We will estimate and make valid conclusions about continuous outcome measure(s):
  for single populations/group, two or more populations/groups.

```

## Hypothesis testing for numerical data

### a) One group

```{r, eval=FALSE}

one sample - One Sample T-test

E.g., Mean weight (Kg) of RVCOP class participants

```

### b) Two (2) groups

```{r eval=FALSE}

i) Independent groups

E.g., Mean weight of boys and girls in a class

Test: Independent 2 sample T-test

ii) Paired groups/matched observations

E.g., Mean weight of boys before and after special diet in a class

Test: Paired 2 sample T-test

```

### Three (3) or more groups

```{r eval=FALSE}

i) Independent groups

E.g., Mean glucose levels of men, women and children in a village

H0: ua = ub = uc // ud = 0
HA: ud != 0

Test: One-way ANOVA - (ANalysis Of VAriance)

ii) Paired groups/matched observations

E.g., Mean glucose levels of diabetic patients one hour after 3 drugs experiments

Test: Repeated Measures ANOVA

```

## 1) a) One group - One sample t-test 

```{r eval=FALSE}

Make inference about a continuous outcome in single population; i.e. 
to determine/compare a sample mean with specified population mean (hypothesized value).

The sample mean is thus used to make decision about the hypothesized population mean

Define the null hypothesis (H0), the alternate hypothesis (H1) and the hypothesized population mean value

If the p-value is less than chosen level of significance (p<0.05), we reject null hypothesis and report that:
  
  “the data provides statistical evidence to indicate that the 
alternative hypothesis is true”. 

If the p-value is equal to or greater than level of significance (p≥0.05), we fail to reject the null hypothesis and report that:
  
  “the data does not provide sufficient evidence to indicate that the alternative hypothesis is true”

```


```{r eval=FALSE}

> t.test(data, alternative, mu, conf.level)

> data: data on the variable of interest
> alternative: what type of alternative hypothesis is specified?   (options: “two.sided”, “greater”, “less”)
> mu: the value of μ under the null hypothesis
> conf.level: confidence level of the test (1-α)

```

```{r}

# define a vector/data of orange weights
orange_weights <- c(300, 315, 320, 311, 314, 309, 300, 308, 305, 303, 305, 301, 303)

t.test(orange_weights, mu = 310, conf.level = 0.95)

```

## 2) a) 2 Independent Groups - Two sample t-test 

```{r eval=FALSE}

TWO-SAMPLE t-TEST

t.test(data1, data2, alternative, mu, conf.level)

> data1: data on the first variable of interest
> data2: data on the second variable of interest
> alternative: what type of alternative hypothesis is specified?   (options: “two.sided”, “greater”, “less”)
> mu: the value of the difference μ−μ2 under the null hypothesis
> conf.level: confidence level of the test (1−α)

```


```{r}
data1 <- c(300, 315, 320, 311, 314, 309, 300, 308, 305, 303, 305, 301, 303)
data2 <- c(335, 329, 322, 321, 324, 319, 304, 308, 305, 311, 307, 300, 305)

t.test(data1, data2) # alternative, mu, conf.level)

iris <- iris %>% filter(!Species %in% c('versicolor'))

boxplot(Sepal.Length ~ Species, data = iris)

t.test(Sepal.Length ~ Species, data = iris)


```

## 3) a) 2 groups paired - Two sample paired t-test 

```{r, echo=FALSE}

#define before and after max jump heights
before <- c(22, 24, 20, 19, 19, 20, 22, 25, 24, 23, 22, 21)
after <- c(23, 25, 20, 24, 18, 22, 23, 28, 24, 25, 24, 20)

t.test(x = before, y = after, paired = TRUE)


```

## 4) a) Three (3) or more independent groups - One way ANOVA

```{r eval=FALSE}

ANOVA - a generalization of the t-tests.

This ANOVA test is an extension of the two independent samples t-test.

It is a test for equality of three or more populations means 
for independent groups and uses sample means of respective populations to draw conclusion.

E.g., A JAVA stores is test marketing of 3 pizzas in Kenya. To find out if they are of the same popularity, 18 joints are randomly chosen for participation in the study. 6 of 18 joints are randomly chosen to test market the first new pizza, another 6 for the second, and the remaining 6 for the last pizza.

piz1 piz2 piz3 
22    52    16 
42    33    24 
44     8    19 
52    47    18 
45    43    34 
37    32    39

i) Assumptions: Outcome measurements - normally distributed
ii) The population variance is the same across the groups (homogeinity of variance). #Bartlett's test

Hypotheses:
  
H0 - all the populations have equal means/
Population means are equal for all groups - any observed differences in sample means are due to random sampling variation.

H1 - at least two population have different mean values/
Observed differences between sample means are due to actual differences in the population means.

**Variance** is used to assess equality of the mean values of different groups.

```


```{r}
pizza <- data.frame(
  piz1 = c(22, 42, 44, 52, 45, 37),
  piz2 = c(52, 33, 8, 47, 43, 32),
  piz3 = c(16, 24, 19, 18, 34, 39)
)

library(tidyr)
long_pizza <- pizza %>% pivot_longer(., cols = c(piz1, piz2, piz3),
                                     names_to = 'Type',
                                     values_to = 'Sales')

summary(aov(Sales ~ Type, data = long_pizza))


```

## 5) b) Three (3) or more paired/matched groups - Repeated Measures ANOVA Test


```{r eval=FALSE}

Repeated measures design is a research design that involves
multiple measures of the same variable taken on the same or matched subjects either under different conditions or over two or more time periods. 

E.g., in a longitudinal study, measurements of glucose levels are taken from the same diabetic patients at different time points in which change over time is assessed following treatments.

In this case, the sample values are correlated - this test is therefore  ANOVA for correlated samples. It is a generalization of paired T-tests.

Assumptions:
  i) Response measurements are normally distributed.
  ii) Covariance between responses in each subject are equal (sphericity)  

*Use Mauchly`s* Test for Sphericity* to evaluate equality of variances of the difference between all possible pairs of repeated condition.

install.packages('rstatix')
library(rstatix)
library(stats)

```

```{r, echo=FALSE, message=FALSE}

diabetes_data <- data.frame(
subject = c(rep("Subject 1", 3), rep("Subject 2", 3), rep("Subject 3", 3)),
time = rep(c("Pre", "Post", "Follow-up"), 3),
score = c(10, 12, 15, 8, 9, 11, 13, 14, 16),
id = c(rep(c(1), 3), rep(c(2), 3), rep(c(3), 3))
)

library(ggpubr)

ggpubr::ggqqplot(diabetes_data, "score", facet.by = "time")

```

```{r}

ggpubr::ggqqplot(diabetes_data$score)

```

```{r}

shapiro.test(diabetes_data$score)

diabetes_data %>% group_by(time) %>%  rstatix::shapiro_test(score)

```

```{r}

library(stats)
library(car)

rep_anova <- aov(score ~ time + Error(subject/time), data = diabetes_data)

model <- aov(score ~ subject, data = diabetes_data) # needs for sphericity correction

summary(model, sphericity.correction = "GGe") # sphericity correction applied

#mauchly.test(model)
```

Perform repeated measures ANOVA
Run repeated measures ANOVA

```{r}
library(ez)

#?ezANOVA # read documentation of the function

anova_result <- ezANOVA(
  data = diabetes_data,
  dv = score,
  wid = id,
  within = time,
  between = subject, #measurements at different times within same patients
  detailed = TRUE,
  type = 3)

anova_result

```

Repeated measures ANOVA with Sphericity correction

```{r}
library(afex)

anova_result <- aov_ez(
  data = diabetes_data,
  id = "id",
  dv = "score",
  within = "time",
  between = "subject",
  return_aov = TRUE,
  anova_table = list(correction = "GG")
)

anova_result

```


