---
title: "Ordinal Logistic Regression Model in R"
author: "John Wainaina <phomwangi@gmail.com>"
date: "2023-06-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


```

## Ordinal Logistic Regression Model

**Introduction**

The model estimates the relationship between the independent variables and the odds of an observation falling into a higher category relative to a lower category.

The key assumption - proportional odds assumption!

 ... states that the relationship between the independent variables and the dependent variable is consistent across all levels of the dependent variable. 
 
 ... means; the effect of the independent variables on the odds of being in a higher category versus a lower category is assumed to be constant.


```{r}

# Fit an ordinal logistic regression model

library(MASS)
library(foreign)
library(haven)
library(mlbench) 
library(ordinal)

data(wine) 

summary(wine)

model <- clm(rating ~ temp  +  contact, data = wine)

summary(model)

coef(model) # Threshold coefficients and predictor estimates:


```
Threshold coefficients represent the cutoff points or boundaries between different rating categories. 

They indicate the point at which the log-odds of the response variable transitioning from one category to the next changes.


# Interpret coefficient estimates

***tempwarm***: The coefficient estimate of 2.5031 suggests that when the temperature is warm, there is an increase in the log-odds of a higher rating for the wine. The coefficient is statistically significant (p < 0.001), indicating a strong positive effect of temperature on the wine rating.

the coefficient of 2.5031 means that when the temperature changes from the reference level (e.g., cool or cold) to warm while holding other variables constant, the log-odds of a higher rating increase by a factor of exp(2.5031) or approximately 12.23. This suggests a substantial positive effect of warm temperature on the wine rating.


***contactyes***: The coefficient estimate of 1.5278 indicates that when there is contact, there is an increase in the log-odds of a higher rating for the wine. The coefficient is statistically significant (p = 0.00135), suggesting a significant positive effect of contact on the wine rating.

the odds of a higher rating for the wine increase by a factor of exp(1.5278) or approximately 4.61. This means that wines with contact have about 4.61 times higher odds of receiving a higher rating compared to wines without contact, while holding other variables constant.


## Convert the coefficients to Odd Ratios


```{r}

exp(coef(model))

exp(confint(model))

```
**Interpreting these odds ratios**:

For tempwarm, the odds of a higher rating for the wine are approximately 12.23 times higher when the temperature is warm compared to cool or cold, while keeping other factors constant.

For contactyes, wines with contact have approximately 4.61 times higher odds of receiving a higher rating compared to wines without contact, assuming other variables remain the same.

These odds ratios provide an understanding of the relative change in the odds of a higher rating associated with each predictor variable.

**Inteprete thresholds with ORs**:

Threshold 1|2: The exponentiated coefficient of -1.3444 corresponds to an odds ratio of exp(-1.3444) ≈ 0.261. This means that, compared to the reference level (category 1), the odds of wines being classified into rating category 2 or higher are approximately 0.261 times lower. The odds of wines being in a higher rating category (2 or above) decrease by about 74% compared to category 1.

Threshold 2|3: The exponentiated coefficient of 1.2508 corresponds to an odds ratio of exp(1.2508) ≈ 3.49. This indicates that, compared to rating category 2, the odds of wines being classified into category 3 or higher are approximately 3.49 times higher. Wines are about 3.49 times more likely to be in a higher rating category (3 or above) compared to category 2.


```{r}
# Example datasets for ordinal



```

```

